Cooling Requirements for the Vertical Shear Instability in astrophysical disks \\ and Implications for Protoplanetary Disks

### 2015-04-09 appendix.tex
restructed, merged with youdin's abstract

\subsection{Spurious growth of adiabatic perturbations in the VGSB}\label{analytic_adia} 
A limitation of the VGSB is that it cannot be used in the
adiabatic limit, even if the above conditions are satisfied. We explain this by considering vertically isothermal
disks and setting  $\beta\to\infty$ in 
Eq. \ref{vertiso_gov_nondim} % (which itself is obtained by setting
% $\hat{g}_c=0$ in Eq. \ref{lin_mass}---\ref{lin_energy})
to give 
\begin{align}
  0 =\dd v_z^{\prime\prime} + \left(1 + \ii \epsilon q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime
  +\left\{\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) 
    -\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii\epsilon  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}
We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  \hat{\sigma}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat 
  =&  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
&+
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii\epsilon
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat
+ \ii\epsilon q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
% When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
% $\hat{\sigma}^2$ is real, and  $\hat{\sigma}^2>0$ if $\gamma>1$. As
% expected, a sub-adiabatically stratified disk is stable in the absence
% of vertical shear. 
In the presence of vertical shear $q\neq0$, Eq. \ref{adia_integral}
suggests $\hat{\sigma}^2$ is complex for real $\khat$, which implies the
possibility of instability, regardless of the value of
$|\gamma-1|$. This is contradicts the second Solberg-Hoiland criterion 
(Eq. \ref{solberg2}), which imply instability for adiabatic
perturbations in a thin, vertically isothermal disk would require 
$|\gamma-1|\ll1$. 
%also need real freq^2 in adia limit 
This is because the VGSB retains the vertical shear
($\p_z\Omega$) from the global disk, but ignores the global radial
disk structure responsible for it. Nevertheless, we demonstrate below
that this inconsistency is unimportant for the VSI because it occurs
for $\beta\ll1$.   %and small scale

@@

\section{Linear problem in the isothermal limit}\label{iso_discuss}  
Here we summarize selected results for isothermal
perturbations ($\beta\equiv 0$) in vertically isothermal disks 
($\Gamma=1$) in the VGSB framework. In this case Eq. \ref{ode_Q}
becomes    
\begin{align} 
  Q = W 
\end{align}
(i.e. $\delta P = c_s^2\delta \rho$), and so $  \ii\hat{\sigma}c_s\delta v_z = W^\prime $
from Eq. \ref{lin_vz}. For isothermal perturbations it is 
simpler to work with an equation for $W$ by substituting $Q=W$ into
Eq. \ref{ode_w}. In this case, we shall not yet make the low frequency
approximation, but first make the  Keplerian approximation. We
obtain, in terms of dimensionless variables, 
\begin{align}
  0 = W^{\prime\prime} + \left[\ln\rho^\prime - \frac{\ii \epsilon q \hat{k} f(\zhat)
      }{1-\hat{\sigma}^2} \right]W^\prime +
  \hat{\sigma}^2\left(1+\frac{\hat{k}^2}{1-\hat{\sigma}^2}\right)W,\label{iso_ode0} 
\end{align}
where for discussion purposes we have defined $f(\zhat)$ such that
\begin{align}\label{fz_shear}
  \frac{d\Omega^2}{d\hat{z}} = \epsilon^2q f(\hat{z})\Omega_k^2.
\end{align}
By comparing Eq. \ref{fz_shear} with Eq. \ref{vertical_shear_ex}, we
see that $f(\zhat)= 
\zhat\left(1+\epsilon^2\zhat^2\right)^{-3/2}$. More generally, though,
$f$ may be regarded as a representation of the vertical
shear profile. Physically, we expect there is a maximum value of 
$|d\Omega/dz|$, the existence of which should limit the growth rate of
the VSI,  as remarked in \S\ref{vshear_def}. We explicitly demonstrate
this below.

>>>>

\subsection{Spurious growth of adiabatic perturbations in the VGSB}\label{analytic_adia} 
A limitation of the VGSB is that it cannot be used for 
adiabatic flow, even if the above conditions are satisfied. We explain this by considering vertically isothermal
disks and setting  $\beta\to\infty$ in 
Eq. \ref{vertiso_gov_nondim} % (which itself is obtained by setting
% $\hat{g}_c=0$ in Eq. \ref{lin_mass}---\ref{lin_energy})
to give 
\begin{align}
  0 =\dd v_z^{\prime\prime} + \left(1 + \ii \epsilon q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime
  +\left\{\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) 
    -\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii\epsilon  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}
We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  \hat{\sigma}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat 
  =&  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
&+
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii\epsilon
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat
+ \ii\epsilon q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
% When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
% $\hat{\sigma}^2$ is real, and  $\hat{\sigma}^2>0$ if $\gamma>1$. As
% expected, a sub-adiabatically stratified disk is stable in the absence
% of vertical shear. 
In the presence of vertical shear $q\neq0$, Eq. \ref{adia_integral}
suggests $\hat{\sigma}^2$ is complex for real $\khat$, which implies the
possibility of instability, regardless of the value of
$|\gamma-1|$. This is contradicts the second Solberg-Hoiland criterion 
(Eq. \ref{solberg2}), which imply instability for adiabatic
perturbations in a thin, vertically isothermal disk would require 
$|\gamma-1|\ll1$. 
%also need real freq^2 in adia limit 
This is because the VGSB retains the global radial
disk structure responsible for vertical shear
($\p_z\Omega$), but ignores it elsewhere in the linear problem. 
Nevertheless, we demonstrate below
that this inconsistency is unimportant for the VSI because it occurs
for $\beta\ll1$.   %and small scale

@@

\section{Linear problem in the isothermal limit}\label{iso_discuss}  
Here we summarize selected results for isothermal
perturbations ($\beta\equiv 0$) in vertically isothermal disks 
($\Gamma=1$) in the VGSB framework. In this case Eq. \ref{ode_Q}
becomes    
\begin{align} 
  Q = W 
\end{align}
(i.e. $\delta P = c_s^2\delta \rho$)% , and so $  \ii\hat{\sigma}c_s\delta v_z = W^\prime $
% from Eq. \ref{lin_vz}
. For isothermal perturbations it is 
simpler to work with an equation for $W$ by substituting $Q=W$ into
Eq. \ref{ode_w} and using Eq. \ref{ode_vz} to eliminate $\delta v_z$. 
In this case, we shall not yet make the low frequency
approximation, but first make the  Keplerian approximation. We
obtain, in terms of dimensionless variables, 
\begin{align}
  0 = W^{\prime\prime} + \left[\ln\rho^\prime - \frac{\ii \epsilon q \hat{k} f(\zhat)
      }{1-\hat{\sigma}^2} \right]W^\prime +
  \hat{\sigma}^2\left(1+\frac{\hat{k}^2}{1-\hat{\sigma}^2}\right)W,\label{iso_ode0} 
\end{align}
where for discussion purposes we have defined $f(\zhat)$ such that
\begin{align}\label{fz_shear}
  \frac{d\Omega^2}{d\hat{z}} = \epsilon^2q f(\hat{z})\Omega_k^2.
\end{align}
By comparing Eq. \ref{fz_shear} with Eq. \ref{vertical_shear_ex}, we
see that $f(\zhat)= 
\zhat\left(1+\epsilon^2\zhat^2\right)^{-3/2}$. More generally, though,
$f$ may be regarded as a representation of the vertical
shear profile. Physically, we expect there is a maximum value of 
$|d\Omega/dz|$, the existence of which should limit the growth rate of
the VSI,  as remarked in \S\ref{vshear_def}. We explicitly demonstrate
this below.

### 2015-04-09 paper.tex
restructed, merged with youdin's abstract

\title{Cooling Requirements for the Vertical Shear Instability, \\
 with Implications for Protoplanetary Disks}
%\title{Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula} 
\author{Min-Kai Lin \& Andrew Youdin}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\email{minkailin@email.arizona.edu, youdin@email.arizona.edu}

\begin{abstract}
  It is difficult to understand how cold circumstellar disks accrete onto their central stars.  Due to 
  lower levels of ionization, the magnetic transport mechanisms that operate in warmer disks are less effective. 
  A hydrodynamic mechanism, the vertical shear instability (VSI), offers a means to drive angular momentum 
  transport and accretion flows in cold protoplanetary disks.  
  The VSI is driven by a weak vertical gradient in the disk's orbital motion.
  In order to grow, the VSI must overcome vertical buoyancy, a strong stabilizing influence in cold disks, where 
  heating is dominated by external irradiation.  Rapid cooling, via radiative losses, reduces the effective buoyancy
  and allows the VSI to operate.  In this paper, we quantify the cooling timescale, $t_c$, needed for growth of the VSI
  and consider the implications for protoplanetary disks where $t_c$ is determined by the opacity of dust 
  grains.  We perform a linear analysis of the VSI with cooling in vertically global and radially local disk models.
  For irradiated disks, we find that the VSI is most vigorous for rapid cooling with $t_c < \Omega_K^{-1} \epsilon |q| / (\gamma -1)$
  in terms of the Keplerian orbital frequency, $\OmK$; the disk's aspect ratio, $\epsilon \ll 1$; the radial 
  powerlaw temperature gradient, $q$; and the adiabatic index, $\gamma$.  For longer cooling times, the VSI 
  is much less effective because growth slows and shifts to smaller length scales, which are more prone to 
  viscous or turbulent decay.  We apply our results to a  Minimum Mass Solar Nebula (MMSN) disk model.  We
  find that the VSI is most effective at intermediate radii, from $\sim 5$AU to $\sim 50$AU with a
  characteristic growth time of $\sim 30$ local orbital periods.  Growth is suppressed by long cooling times both in the opaque 
  inner disk and the optically thin outer disk.  A reduction in the dust opacity by a factor of 10 increases cooling times 
  enough to quench the VSI at all disk radii.  Thus the formation of solid protoplanets, a sink for dust grains, can impede the VSI.

>>>>

\title{Cooling Requirements for the Vertical Shear Instability in
  astrophysical disks \\ and Implications for Protoplanetary Disks}
%\title{Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula} 
\author{Min-Kai Lin \& Andrew Youdin}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\email{minkailin@email.arizona.edu, youdin@email.arizona.edu}

\begin{abstract}
  It is difficult to understand how cold circumstellar disks accrete onto their central stars.  Due to 
  lower levels of ionization, the magnetic transport mechanisms that operate in warmer disks are less effective. 
  A hydrodynamic mechanism, the vertical shear instability (VSI), offers a means to drive angular momentum 
  transport in cold accretion disks such as protoplanetary disks.  
  The VSI is driven by a weak vertical gradient in the disk's orbital motion.
  In order to grow, the VSI must overcome vertical buoyancy, a strong stabilizing influence in cold disks, where 
  heating is dominated by external irradiation.  Rapid cooling, via radiative losses, reduces the effective buoyancy
  and allows the VSI to operate.  In this paper, we quantify the cooling timescale, $t_c$, needed for growth of the VSI
  and consider the implications for protoplanetary disks where $t_c$ is determined by the opacity of dust 
  grains.  We perform a linear analysis of the VSI with cooling in vertically global and radially local disk models.
  For irradiated disks, we find that the VSI is most vigorous for rapid cooling with $t_c < \Omega_K^{-1} \epsilon |q| / (\gamma -1)$
  in terms of the Keplerian orbital frequency, $\OmK$; the disk's aspect ratio, $\epsilon \ll 1$; the radial 
  powerlaw temperature gradient, $q$; and the adiabatic index, $\gamma$.  For longer cooling times, the VSI 
  is much less effective because growth slows and shifts to smaller length scales, which are more prone to 
  viscous or turbulent decay.  We apply our results to a  Minimum Mass Solar Nebula (MMSN) disk model.  We
  find that the VSI is most effective at intermediate radii, from $\sim 5$AU to $\sim 50$AU with a
  characteristic growth time of $\sim 30$ local orbital periods.  Growth is suppressed by long cooling times both in the opaque 
  inner disk and the optically thin outer disk.  A reduction in the dust opacity by a factor of 10 increases cooling times 
  enough to quench the VSI at all disk radii.  Thus the formation of solid protoplanets, a sink for dust grains, can impede the VSI.
